---
title: 'Psych 252: R Tutorial Module 2'
output:
  html_document:
    highlight: pygments
    theme: flatly
    toc: yes
---

Based on notes by Paul Thibodeau (2009) and revisions by the Psych 252 instructors in 2010 and 2011

Expanded in 2012 by Mike Frank, Benoit Monin and Ewart Thomas

Converted to [R Markdown](http://www.rstudio.com/ide/docs/r_markdown) format and further expanded in 2013 by Steph Gagnon.



```{r detach_rtw_2}
detach(rtw.table)
```

Plotting Data from Repeated Measures Designs: An intro to ggplot2 & reshape2
----------------------------------------------------------------------------

[ggplot2](http://ggplot2.org/) is a plotting library for R. It is an easy way to create beautiful plots. There are a *ton* of different options, so feel free to explore these with Google as a guide. [Here](http://www.cookbook-r.com/Graphs/) is just one of many helpful websites.

### Installing packages

R has a lot of great [packages](http://cran.r-project.org/web/packages/available_packages_by_name.html) contributed by other people who use R. These packages can be really helpful, especially for plotting, and running more complicated analyses.

Let's install a couple useful packages, including **ggplot2** and **reshape2**. [Reshape2](http://stats.stackexchange.com/questions/7439/how-to-change-data-between-wide-and-long-formats-in-r) will allow us to reshape dataframes, e.g., going from short-form (where each participant is in 1 row, with multiple columns for different measurements) to long-form (where each observation is in a different row).

```{r install_ggplot2}
# uncomment this to run! ‚???'‚O~C
#install.packages("ggplot2")
#install.packages("reshape2")
```
If you have trouble with the CRAN mirror, you might need to install using the GUI, by clicking on Tools, and then Install Packages. 



Basically, we now have a long-form dataframe with 3 rows for each subject.
