// This is a file full of unused code, ripped from the old tutorials

/// removed from tutorial 1:
 Additionally, R does not care too much about the distinction between integers and floating point numbers, and in fact most things are floats behind the scenes unless you go out of your way to force integer representation.

///

### Matrices

Matrices are a lot like vectors, except they have two dimensions. You can create them in a few ways. The `matrix` function takes a vector and reshapes it into a matrix based on the other arguments:

```{r matrix_function}
v = 1:6
matrix(v, nrow=2)
matrix(v, ncol=3)
matrix(v, nrow=2, byrow=TRUE)
```

You can also glue several vectors together into a matrix either along rows or along columns:

```{r matrix_bind}
v_1 = 1:6
v_2 = 7:12
rbind(v_1, v_2)
cbind(v_1, v_2)
```

Matrix elements can also have names. As you see above, the `rbind` and `cbind` functions set the names of the rows/columns to the variable names that were passed into them. Alternatively, the `rownames` and `colnames` functions can be used much like the `names` function for vectors:

```{r assign_matrix_names}
m = matrix(1:6, nrow=2)
rownames(m) = c("a", "b")
colnames(m) = c("x", "y", "z")
m
```

Some functions can compute descriptive statistics to collapse one of the dimensions in the matrix and return a vector:

```{r matrix_dim_funcs}
rowMeans(m)
colSums(m)
```

You can also index into a matrix to expose either a single element or an entire row/column:

```{r matrix_index}
m[1, 2]
m[1,]
m[,3]
m["a", "z"]
m["b",]
m[,"y"]
m["a",]["x"]
```

The rules about datatypes also apply to matrices.



//// from module 2


Attaching Dataframes
--------------------

Sometimes it can get annoying typing in the name of the dataframe each time you want to specify a variable. If you're going to be working with one dataframe for a while, it can be helpful to **attach** the dataframe while you're working, and then **detach** it when you're done. When a dataframe is attached, you can just type in the name of the variable, and R will know to look in the attached dataframe.

*Keep in mind that while this can be convenient for interactive analyses, it can be a source of tricky bugs (where names overlap) and should be used with caution and discretion.* 

```{r plot_attached_df}
attach(rtw.table)
plot(URate, COL, pch = 23, bg = 'red', main = 'Cost of living vs. Uneployment Rate')
lines(lowess(URate, COL), lty = 2, col='red')
```

Make sure you remember to detach the dataframe when you're done, otherwise R can get confused if you work with a different dataframe with similar variables!

```{r detach_rtw}
detach(rtw.table)
```


///

### Loading packages
You only need to install these packages once, but you'll need to load them in everytime you start R using the `library()` function. The distinction is that `install.packages()` downloads the files from CRAN and places them in a location where R can find them, while `library()` makes it such that you can call all of the functions that are defined in these packages.

```{r loadlibs_ggplot}
library(ggplot2)
library(reshape2)
```

///






////

